Modeling Hereditary Diffuse Leukoencephalopathy with Axonal Spheroids using microglia-sufficient brain organoids

Version 1

Abstract

Abstract

Hereditary diffuse leukoencephalopathy with axonal spheroids (HDLS) is a rare, fatal, adult-onset neurodegenerative disease that is most often caused by mutations affecting the Colony Stimulating factor-1 Receptor (CSF-1R). To understand how CSF-1R-mutation affects human microglia – the specialized brain-resident macrophages of the central nervous system – and the downstream consequences for neuronal cells, we used a macrophage and forebrain organoid co-culture system based on induced pluripotent stem cells generated from two patients with HDLS, with CSF-1R gene-corrected isogenic organoids as controls. Macrophages derived from iPSC (iMacs) of patients exhibited a metabolic shift towards the glycolytic pathway and reduced CSF-1 sensitivity, which was associated with higher levels of IL-1β production and an activated inflammatory phenotype. Single-cell RNA sequencing revealed that iMacs adopt a reactive state that leads to impaired regulation of neuronal cell populations in organoid cultures, thereby identifying microglial dysregulation and specifically IL-1β production as key contributors to the degenerative neuro-environment in HDLS.

Introduction

Hereditary diffused leukoencephalopathy with axonal spheroids (HDLS) is a rare, inherited, autosomal dominant, neurodegenerative disease that is characterized by patchy axonal swellings (spheroids) and demyelination that result in the alteration of the white matter of the brain (1). Although detailed epidemiological data are lacking, in 2021, it was estimated that approximately a quarter of a million people globally are affected by the condition (2); yet our understanding of the pathogenesis of the disease is limited, and there is no cure or even effective treatment.

For more than a decade, we have known that HDLS is caused by mutations affecting the colony-stimulating factor-1 receptor (CSF-1R) (3) which most often occur in the tyrosine kinase domain (TKD) (4), and lead to reduced auto-phosphorylation in response to CSF-1, which results in impaired downstream signaling (4, 5). These mutations most profoundly affect microglia (6), the specialized brain-resident macrophages that play critical roles in brain development and homeostasis (7, 8) as well as in a wide range of neuroinflammatory and neurodegenerative diseases (9). Patients with HDLS have few, small, microglia that are distributed abnormally within the neocortex and that express low levels of the homeostatic marker P2RY12 and high levels of the inflammatory marker CD68, compared to healthy controls (10–12).

Studies aiming to understand the pathophysiology of HDLS have used various approaches. Rodent models have been generated, but do not faithfully recapitulate the disease phenotype in humans: the earliest haplo-insufficient Csf1r+/- mice model exhibited microgliosis within brain regions (13), but this was not recapitulated in a later Csf1r knockout rat model (14); while a subsequent orthologous mouse model has phenocopied the reduced microglia density seen in patients with HDLS but does not exhibit brain pathology (15). A zebrafish model of HDLS similarly saw reduced microglia during brain development (16), but does not offer the opportunity to study pathology associated with adult-onset neurodegenerative diseases (17). To overcome these species-specific limitations, in this study we generated human induced pluripotent stem cell (iPSC) - derived macrophages from patients with HDLS and used them to create microglia-sufficient autologous brain organoids (18) in which we characterized the effects of naturally occurring CSF-1R mutation on microglial function. By comparing iPSC-derived macrophages to genetically corrected macrophages from the same patients, we uncovered evidence of significant transcriptional and metabolic reprogramming that was associated with profound changes to neuronal regulation by these cells, indicative of the likely mechanistic links between microglial dysfunction and HDLS.

Results

Generation and characterization of iPSC lines from patients with HDLS

We first isolated dermal fibroblasts from skin biopsies taken from two patients with HDLS and confirmed mutations in the TDK region of the CSF-1R gene (HD1, HD2; clinical characteristics shown in Supplemental Table 1) then cultured them for approximately four weeks before reprogramming into induced pluripotent stem cells (iPSCs) (Fig 1A). Four to five weeks post-transfection, we observed iPSC-like colonies emerging (Supp Fig 1B), and 7-10 days later, we selected those clones that were proliferating and displayed typical iPSC morphology (Fig 1C). We confirmed successful reprogramming by visualization of pluripotency markers including SOX2 and OCT4 (nuclear markers), and TRA-1-60 and SSEA-4 (intracellular marker), by immunofluorescence (Fig 1D). We also confirmed that these clones had normal karyotype (Fig 1E) and termed them Mut HD1 and Mut HD2. One clone per cell line was used for further analysis. We then generated isogenic from Mut HD1 and Mut HD2 iPSC cell lines which we termed IsoHD1 and IsoHD2 respectively, as controls (see Methods). Clones that were karyotypically normal were used for downstream experiments alongside their mutant counterparts (Fig 1B), with no off-target events identified (Supp Fig 1C). Collectively, HDLS iPSC and their corresponding isogenic iPSC controls were successfully generated from two patients with HDLS (HD1 and HD2).

Generation and characterization of iPSC lines from patients with HDLS

A) Schematic representation of the generation of patient-derived induced pluripotent stem cells (iPSC) from dermal fibroblasts, and the subsequent generation of isogenic controls via gene editing with CRISPR/Cas9. B) Sequencing of the genomic CSF-1R locus in donor-derived iPSC displaying mutant allele or genetically edited allele in isogenic control cell lines. C) Phase-contrast microscopy of mature reprogrammed iPSC colonies displaying typical hESC morphology. Scale bar represents 250 uM. D) Immunofluorescence imaging of iPSC colonies for the pluripotency markers SOX2, OCT4, TRA-1-60 and SSEA-4. Scale bar represents 150 uM and 100 uM for left and right panel respectively. E) Cytogenetic analysis of mutant and genetically edited iPSC clones showing a normal karyotype.

A) Human dermal fibroblast outgrowth from skin biopsies after three weeks of culture. B) Phase-contrast microscopy of emerging immature iPSC clones after reprogramming. C) Sequencing of genomic CSF1R locus displaying off-target events (OTE) analysis for IsoHD1 and IsoHD2 iPSC cell lines.

Generation and characterization of patient iPSC-derived macrophages

We next sought to differentiate primitive microglia-like macrophages from patient-derived iPSC cell lines (iMacs) using our serum-free protocol that faithfully recapitulates the generation of in vivo microglial progenitor cells (19). Briefly, iPSCs were initially inducted towards the mesodermal lineage before being differentiated into hemangioblast-like cells with the potential to differentiate into both endothelial and hematopoietic progenitors - the latter emerging as free-floating cells as early as day six after seeding - which we collected and then terminally differentiated into iMacs by the addition of CSF-1 (Fig 2A). Although HDLS mutation downregulates the sensitivity of CSF-1R to its ligand (5), we were able to successfully generate iMacs using this protocol: the yield of iMacs from iPSC varied between cultures (Fig 2B), but flow cytometry confirmed their consistent macrophage-like phenotype, with abundant expression of CD45 and CD14 (Fig 2C), and Giemsa staining showed that iMacs exhibited comparable morphology between donors (Fig 2D). Functionally, all iMac cultures were equally able to take up PE-labelled beads that were detectable within their cell bodies (Fig 2D); flow cytometry confirmed that the level of actin-dependent phagocytosis of pHrodo beads was comparable among CD45-expressing cells from all cultures (Fig 2E). Collectively, these data show that iMacs generated from HDLS donor-derived iPSC adopted the typical morphology, phenotype and functions associated with macrophages.

Generation and characterization of patient iPSC-derived macrophages

A) Schematic representation of the derivation of primitive macrophages from donor-derived iPSC. B) iMacrophage (iMac) yields across all cell line variants. C) Flow cytometry analysis of iPSC and iMac expression of macrophage surface markers CD45 and CD14. D) Morphology of iMac visualized by Giemsa staining (left panel), and confocal imaging (middle panel); and internalization of PE-conjugated beads visualized by confocal imaging. Scale bar represents 30 uM. E) Uptake of pHrodo-labeled beads by iMac in the presence and absence of cytochalasin D (Cyto D). F) Cell viability of iMacs used to determine CSF1 sensitivity via an MTT assay, after incubating with varying concentrations of CSF1 for seven days. Data are presented as mean ±SEM.

The mutations associated with HDLS downregulate the sensitivity of CSF-1R to its ligand CSF-1, but to different extents depending on the mutation site (5). We therefore compared the survival of our mutant and isogenic cells lines from both patients during culture with CSF1. As expected, even with high concentrations of CSF-1, relatively few mutant cells survived across seven days of culture; by contrast, even low CSF-1 levels allowed significantly more isogenic cells to survive (Fig 2F).

In summary, we generated iMacs from iPSC of patients with HDLS and their respective isogenic controls that exhibit typical macrophage morphology, phenotype, and phagocytic capabilities.

Mutant and isogenic control iMacs have distinct gene expression profiles

Although iMacs derived from patients with HDLS exhibited core macrophage features, we wanted to more fully understand the impact of CSF-1R mutation on these cells. We first characterized their gene expression profile using bulk RNA-seq and compared it to that of their respective CSF-1R-restored isogenic iMac cultures.

Principal component analysis (PCA) showed that patterns of gene expression differed markedly between mutant and isogenic iMacs. Among cell cultures from HD1, mutant iMacs exhibited patterns of gene expression consistent with a more reactive state compared to isogenic CSF-1R-restored HD1: HD1 Mut had higher expression of genes involved in pathways related to proliferation (IGF2, MYBL2, STAT1, INCEP, TGFA) and the defense response to virus (IFI27, IFI44L, STAT1, MX1, IFIT3); while Iso HD1 iMacs expressed high levels of genes involved in chemokine-related responses and signaling pathways (CXCL2, ADAM8, CCL31L, CCL5,), as well as leukocyte migration and chemotaxis (CXCL3, CCL13, CCL2, E5AR1, C3AR1) (Fig3 A-C). We observed a similar transcriptomic profile in HD2-derived iMacs: HD2 Mut showed up-regulated gene expression for the defensive response to virus, while isogenic HD2 iMacs exhibited high expression levels of genes involved in leukocyte migration and chemotaxis (Fig 3D-F). These data are in line with previous reports suggesting that macrophages from patients with HDLS adopt an inflammatory phenotype with increased expression of the immunoreactive markers CD68 and CD163 (10, 12). Taken together, iMac cultures derived from patients with HDLS exhibited transcriptomic profiles consistent with upregulated inflammatory gene expression as a result of impeded CSF-1R function.

Mutant and isogenic control iMacs have distinct gene expression profiles

A and D) PCA analysis, B and E) volcano plots and C and F) GO analysis, of mutant and isogenic iMac derived from HD1 and HD2 differentially expressed genes (DEGs) respectively. Representative curve of ECAR (G, O) and their respective quantified data (H-J, P-R) are shown. Representative curve of OCR (K, S) and their respective quantified data (L-N, T-V) are shown. Real-time assessment of bioenergetic profile between mutant and isogenic HD1 iMac, measured using extracellular flux assay and a mitochondrial stress test. The following glycolytic parameters were calculated based on ECAR: B) glycolysis, C) glycolytic capacity, D) glycolytic reserve. Data shown as means + SEM, n=9. The following parameters were calculated based on OCR: F) basal, G) ATP production and H) maximal respiration. 2-DG = 2-deoxy-glucose, FCCP = carbonyl cyanide 4-trifluoromethoxy-phenylhydrazone. (n=3) Data are presented as mean ± SEM.

Mutant and isogenic control iMacs have distinct metabolic profiles

Studies have shown that activated macrophages often undergo metabolic reprogramming to fulfil the energetic needs of demanding biochemical processes (20, 21); alongside, we know that energy metabolism, as well as metabolic cell features, plays an important role in mediating macrophage function and plasticity (22). Therefore, we next compared the cellular metabolic profiles of patient-derived CSF-1R mutant and isogenic CSF-1R-restored iMacs.

We first assessed glycolysis by measuring the extracellular acidification rate (ECAR) in the presence of specific modulators (glucose, oligomycin, 2-deoxy-glucose) that were added sequentially to the cell cultures to reveal various aspects of the metabolic pathway (see Methods): this showed that CSF-1R-mutant iMacs from both HD1 and HD2 had a significant overall increase in glycolytic function (Fig 3G and 3O). Although the magnitude of the difference between iMac HD1 and HD2 and their respective CSF-1R-restored counterparts varied, in both cases, CSF-1R mutation was associated with significantly increased glycolytic capacity and glycolytic reserve.

Next, we determined the functional metabolic profile of mitochondrial respiration by measuring real-time changes in the extracellular oxygen consumption rate (OCR) during sequential treatment of cells with oligomycin, carbonyl cynide p-trifluoromethoxyphenylhydrazone (FCCP) and a combination of rotenone and antimycin As before, although the magnitude of effect was less for cells from HD1, both sets of iMacs showed a similar trend towards higher OCR values, and greater capacity for basal respiration, ATP production, and maximal respiratory capacity in the presence of mutated CSF-1R: in the case of HD2 Mut cells, these differences achieved statistical significance across all three bioenergetic parameters.

Taken together, these results indicate that the absence of fully functional CSF-1R in cells from HD1 and HD2 is associated with an upregulated glycolytic metabolic profile, which is cells from HD2 is clearly accompanied by a higher mitochondrial oxidative phosphorylation rate. This suggests that macrophages from patients with HDLS have an altered metabolic profile that correlates with their higher activation status and could contribute to disease pathology.

HDLS iMacs upregulate IL-1β when exposed to apoptotic neuronal cells

Our data suggest that mutant iMacs have dysregulated transcriptomic and metabolic profiles, suggestive of an active inflammatory state. To better understand how this would be expressed in the context of their functions in the brain, we mimicked a need for apoptotic cell clearance by incubating iMacs from patients with HDLS with UV-irradiated neuronal SH-SY5Y cells (Supp Fig 2A). Both iMac lines (CSF-1R mutant and restored) from both patients effectively phagocytosed apoptotic but not control, non-UV-irradiated, SH-SY5Y cells, in an actin-dependent manner (Supp Fig 2B).

We then measured the expression of genes encoding pro-(IFN-g, TNF-α, IL-1β, IL-6, IL-18) and anti-(IL-10, IL-12, TGF-β) inflammatory cytokines following incubation of iMacs with UV-irradiated or non-irradiated SH-SY5Y cells, or without any SH-SY5Y cells, for eight hours. We observed a significant and specific increase in IL1B transcription in CSF-1R mutated iMacs from both HD1 and HD2 when exposed to apoptotic cells, compared with their isogenic counterparts. Among the anti-inflammatory cytokines, Mut HD2 cells exhibited significantly increased transcription of the gene encoding TGF-β when cocultured with UV-treated cells, alongside high baseline levels of IL-12 and TGF-β in control cultures compared to isogenic HD2 iMacs. Interestingly, all CSF-1R mutant iMac groups had higher levels of transcription of most pro-inflammatory cytokines at baseline, consistent with their activated/reactive transcriptional profile, than did their CSF-1R-restored counterparts (Fig 4A-B).

Cytokine profiling of iMacs saw upregulated levels of IL-1β when probed with apoptotic neuronal cells

(A-B) Levels of transcription of pro- and anti-inflammatory cytokine genes (IFNg, TNFa, IL-1b, IL-6, IL-18, IL-10, IL-12 and TGFb) quantified by qPCR. (+uv depicts UV-treated, -uv depicts non-UV treated). n=3 per cell line. Data are presented as mean ±SEM.

A) CFSE-labeled SH-SY5Y cells were either UV-treated (red) or untreated (blue) then after 24 hours were incubated with Annexin V/PE to determine the induction of apoptosis. Annexin V labeling was measured by flow cytometry (top panel) and visualized by fluorescence microscopy (bottom panel). B) Percentage of CD45+ cells in iMac cultures from HD1 (top row) and HD2 (bottom row) that took up apoptotic SH-SY5Y cells, with cytochalasin D used as a negative control. Data are presented as mean ±SEM. Scale bar represents 30 uM.

Collectively, all iMacs exhibited the ability to specifically phagocytose apoptotic cells, but in the absence of a fully functional CSF-1R, this led to high levels of IL1B transcription, which reflected a general propensity towards the transcription of pro-inflammatory cytokine genes, even in the absence of overt stimulation. Importantly, restoration of CSF-1R via gene editing was sufficient to reverse this inflammatory phenotype.

iMacs differentiate into microglia-like cells in a coculture with forebrain organoid

During HDLS, patients typically exhibit degeneration of the frontal lobes of the brain, which contributes to the deficit and decline in cognitive ability, changes to personality and behavior, and dysfunctional motor, social, and language skills (23, 24). Therefore, we next generated iPSC-derived forebrain organoids including autologous iMacs from patients with HDLS to better understand their potential role in the brain.

To better understand how diseased or non-diseased macrophages affect the developing brain, we first generated forebrain organoids from the isogenic cell line variant of the respective donor’s iPSC (see Methods). After 31 days, organoids were harvested (Supp Fig 3A) and co-cultured for seven days with isogenic or mutated iMacs that had been generated in parallel (Fig 5A). At this point, each organoid was removed from its coculture and cultured individually for 16 more days (Fig 5B).

iMac differentiate into microglia-like cells in coculture with forebrain organoids

Schematic diagram of co-culture between Day 31 iMac and forebrain organoid. B) Forebrain organoids depicted before, during and after co-culture with iMacs. Day 31 isogenic forebrain organoids derived from the respective donor iPSC cell line were cocultured with the same isogenic- and mutant-cell-line-derived Day 31 iMacs. 30,000 iMacs were cocultured with each respective organoid derived from HD1 and HD2. Figure shows matured organoids before and after the addition of iMacs during coculture. C) Three-dimensional visualization of co-cultured brain organoids after 23 days. Neurons were labeled for NESTIN and NPCs for SOX2. Microglia-like cells (IBA-1+) were visualized on the surface of organoids. D) Flow cytometry analysis of neurons, NPCs, and macrophage populations within forebrain organoids after 23 days of coculture. E-F) Donor-derived iMacs in forebrain co-cultures undergo distinct changes in gene expression as observed in PCA analysis (E and H), volcano plots (F and I) and GO analysis (G and J) in both HD1 and HD2 variants. Scale bar represents 200 µM.

A) Progression of forebrain organoid development across 30 days. Cells were initially seeded singly before the formation of embryoid bodies (EB) was observed after 2-3 days. Healthy EBs grew and matured, as seen by the formation of distinct neural epithelium in the organoids between Days 12 and 30. Scale bar represents 200 uM. B) Immunofluorescence imaging of neuroepithelial rosette (red, SOX2+) in Day 13 organoid. Developmental cues further supported the formation and expansion of the organoid to develop neuroepithelial tubes or rosettes, which contain predominantly neural stem cells (NPCs) that mimic the developing embryonic human cerebral cortex. C) Three-dimensional visualization of non-co-cultured brain organoids (Day 23). Neurons were labeled for NESTIN (green, cytoplasm) while the NPCs were labeled for SOX2 (red, cytoplasm). Microglia-like cells (IBA1+, white) were not identified, indicating the lack of formation of innately developing microglia. Scale bar represents 200 µM.

We then applied 3D imaging of forebrain organoids to assess their morphology and the distribution/differentiation of the iMacs. This showed that iMacs expressing the microglial marker IBA-1, were mostly located on the surface of the organoids (Fig 5C). We also labelled organoids for NESTIN (a neuronal cell marker) and SOX2 (expressed by neuronal progenitor cells (NPCs). In all of the organoids, neural epithelial rosettes were present, consisting mostly of SOX2+ NPCs (Supp Fig 3B); these structures indicate optimal organoid growth as they demonstrate proper lineage progression leading to forebrain region identity (25) Importantly, organoids to which we did not add iMacs did not contain any IBA-1+ cells (Supp Fig 3C); this was as expected from our previous work (18), but was in contrast to previous studies that reported innately-derived microglia in mesodermal-lineage-containing cerebral organoids (25, 26).

Next, we characterized the different cell populations within organoids after 23 days of coculture and compared iMacs before and after co-culture. After organoid dissociation, cells were labelled with antibodies to differentiate neurons (CD45-/CD184-/CD44-/CD15lo/CD24+), NPCs (CD45-/CD184+/CD271-/CD44-/CD24+) and iMacs (CD45+) before undergoing sorting (Fig 5D) followed by bulk-RNA sequencing. During the course of the work, sufficient iMac numbers for downstream analysis were only generated from coculture with mutant iMacs for HD1 and isogenic iMacs for HD2 respectively. Compared to Day 31 non-co-cultured iMac, HD1 iMacs after co-culture had profoundly altered their gene expression (Fig 5E), with the upregulated DEGs showing an increase in various pathways that are characteristic of microglial functions in the brain such as central nervous system and brain development (CNTN1, ZIC2, EFNA2, GLI3), synapse organization, synapse and neuron organization and development (SLC1A1, DGKB, TRIM67), as well as neurogenesis (CHD11, ADCYAP1) (Fig 5F-G).

Likewise, for HD2 iMacs, PCA also suggested that their patterns of gene expression were markedly different between cocultured and non-cocultured populations (Fig 5H), while GO analysis of the upregulated DEGs revealed an increase in various pathways linked with typical microglial functions in the brain such as glial cell differentiation and gliogenesis (CLU, MDK, CXCR4, RELN, TMEM98, TUBA1A), axonal development (UCHL1, MAP1B, CRABP2, TUBB2B, LGI1) as well as forebrain development (TUBB2B, MDK, CXCR4, RELN, LHX2) (Fig 5I-J). This shows that iMacs undergoing coculture with forebrain organoids have differentiated to a certain extent, into microglia-like cells.

Co-culture results in impaired regulation of neurons within organoids

Microglia not only act as sentinel cells, but are also important contributors to neuroplasticity and neurogenesis in the developing brain that ultimately help shape the brain circuitry (27). Part of this role requires microglia to regulate the population of NPCs and neurons during the unrestrained, highly proliferative phase of early brain development (28, 29); therefore, we next asked whether the microglia-like cells from our iMac-forebrain organoids were also capable of regulating neurogenesis in vitro.

As such, we sought to compare the size of the organoids, as well as the individual major cell population in each organoid, for both variants between isogenic organoids that did not undergo co-culture, and those that underwent 23 days of co-culture by comparing their area before and after co-culture. Co-culture of HD1 saw a significant reduction in the organoid size among both co-culture parameters before and after co-culture. However, comparison of the organoid size on coculture Day 23 only, saw the isogenic control parameter as the only group that had a significant reduction in organoid size when compared to the organoid-only control (Fig 6A). On the contrary, HD2 displayed no significance in organoid size reduction even after co-culture (Fig 6B).

Co-culture results in impaired regulation of neurons within organoids

A) Size of HD1 and HD2 organoids measured before and after co-culture with mutant and isogenic iMacs in both HD1 and HD2 variants. B) The effects of co-culture on neuronal populations in HD2 organoids after 23 days. C) Transcriptomic analysis of co-cultured neurons and NPCs from HD2 organoids after 23 days. Data are presented as mean ±SEM.

Subsequently, these co-cultured organoids were dissociated, stained, sorted and quantified for the respective cell populations. Co-culture of IsoHD1 organoid with IsoHD1 iMac saw a significant reduction in the neuronal population when compared to control organoids that did not underwent co-culture. Conversely, coculture with mutant HD1 iMac saw a smaller degree of neurons reduction but was not significant (Fig 6C). Next, co-culture of IsoHD2 organoid with IsoHD2 iMac saw a slight reduction in neuronal population that was not statistically significant, while the co-culture with mutant HD2 iMac saw no difference when compared with control organoids (Fig 6D). Interestingly, we did not detect any change in the NPC population for either HD1 or HD2 after co-culture, likely because their numbers were relatively low to begin with due to the cellular composition of the forebrain organoid. Similarly, the iMac population was also small in all co-cultures. Transcriptomic analysis of the neurons of HD2 co-culture saw the mutant iMac co-culture having upregulated expression of genes associated with mitochondrial respiration (BCS1L, MRPL18, TSFM, MRPL11, MRPS30) but this phenomenon was not present in the isogenic iMac co-culture (Fig 6E). Conversely, transcriptomic analysis of the NPCs of HD2 co-culture saw the isogenic iMac co-culture having upregulated expression of genes associated with mitochondrial respiration (Fig 6F). Taken together, our model suggests that HDLS mutation affects neuronal population in a microglia-dependent manner and that this phenomenon is variable and mutation-dependent.

Discussion

HDLS is an incurable neurodegenerative disease that begins in adulthood, progresses rapidly and has a grim prognosis. Efforts to elucidate the functional or dysfunctional aspects of the causative mutations in the CSF-1R gene have been hampered by the lack of an accurate disease model that recapitulates the human pathology: although rodent and zebrafish models have been generated and imparted valuable insights towards HDLS (13–16), discrepancies and limitations in each case have led to incomplete recapitulation of the disease’s pathology, specifically in microglia numbers within brain regions and the effects on respective cognitive and sensorimotor tests. Additionally, a model looking into immune-neuronal crosstalk to better understand a neurodegenerative disorder that is primarily associated with microgliopathy, is important to advance the field. Here, we first generated iPSC lines from two patients with HDLS and then used CRISPR/Cas9 technology to reinstate a fully functional CSF-1R gene into some of the lines. While the generation of an HDLS donor-derived iPSC cell line has been reported (30), no other in vitro studies have yet reverted the causative mutation to wild type; a necessary step to understanding the direct impacts of diminished/absent CSF-1R signaling.

We then generated and characterized iMacs - precursors of microglia - from these lines, which exhibited typical morphology, phenotype and phagocytic capacity that were not obviously affected by CSF-1R mutation. Although we were able to generate phenotypically and functionally similar iMac across these lines for example for phagocytic capability, iMac yield varied greatly between the cell lines with a 50% lower yield for Mut HD1 compared to IsoHD1 and, conversely, a higher yield for Mut HD2 200% compared to IsoHD2. This fundamental difference could be attributed to their respective mutations: the nonsense mutation identified in HD2 could not only more profoundly affect receptor signaling, but might also be a major factor during the course of gene editing leading to incomplete reversion to wild type at the epigenetic level, which we speculate was not the case for IsoHD1.

Functionally, while phagocytosis of E.Coli beads was similar across lines, the baseline cytokine production as well as after phagocytosis of apoptotic neuronal cell line was different with an increase of IL-1β secretion in both mutant HD1 and HD2, compared to isogenic HD1 and HD2 iMacs. In the case of phagocytosis of apoptotic cells, this was not due to a difference in phagocytic activity as all cells were able to phagocytose apoptotic cells at a similar level. Importantly, both mutant iMacs have higher levels of baseline pro- and anti-inflammatory cytokine production in the control group without apoptotic cells. This suggest that mutant iMacs are innately in a more reactive or activated state when compared to isogenic iMacs and could reflect the reactive phenotype of HDLS-associated microglia and macrophage identified in vivo (10, 12). Conversely, isogenic iMacs seems to adopt less of the reactive phenotype with the reverted mutation as identified with an overall lesser cytokine production level. Importantly, such reactive phenotype was also highlighted by our transcriptomic and metabolic profiling with upregulated pathways involving pro-inflammatory cytokines and preferential utilization of the glycolytic pathways in the mutant groups, a feature of macrophage activation (31, 32), respectively.

We subsequently incorporated donor-specific and their isogenic controls into a co-culture system using 3D forebrain organoids and iPSC-derived macrophages to better study HDLS. Isogenic and mutant iMacs co-cultured separately with their respective isogenic forebrain organoid differentiate into microglia-like cells with 23 days of coculture as shown by the expression of the microglial marker IBA-1, their ramified morphology as well as their transcriptomic profiles with up-regulation of genes associated with biological pathways that are typical of microglia functions such as neurodevelopment, neurodifferentiation, neurogenesis and brain development.

Subsequently, we analyzed the size of the organoids before and after co-culture as we recently showed that addition of iMac modulated NPC differentiation, limiting their proliferation and promoting axonogenesis, resulting in a reduction of co-culture organoid size (18). Here, while we noticed a significant decreased in the size of isogenic HD1 forebrains co-cultured with isogenic HD1 iMac compared to isogenic HD1 organoids, we did not measure any significant change in organoid size in the isogenic HD1 organoids co-cultured with mutant HD1 iMacs. This suggested that the crosstalk between iMicro and NPC is dysregulated in the context of the HD1 mutation. Since we showed that such size reduction is mediated by cholesterol transfer from iMicro to NPC, further work will be required to characterize cholesterol metabolism of HD iMac.

While we did not observe major changes in cell type composition in these cocultures, transcriptomic analysis of neuronal and NPC population after co-culture showed an increase in mitochondrial respiratory pathways associated genes in the mutant HD2 iMac co-culture neuron and isogenic HD2 iMac co-culture NPCs population respectively. This suggests increased metabolic rates in these respective populations that are contributed to by the different variants of iMac during co-culture, and show that mutated iMicro can change the states of non-mutated isogenic organoid cells. A parallel process in vivo would likely have profound long-term consequences for brain development and homeostasis.

Altogether, we propose that CSF1-R mutation in HDLS may cause subtle yet impactful disruptions to microglia homeostasis due to abnormal metabolic functions in the developing fetal brain, that subsequently compound and result in aberrant microglia functions during adulthood. This could also be mediated by upregulated pro-inflammatory cytokine production such as IL-1β at a basal level or upon clearing of cellular debris during development and adulthood homeostasis. IL1-β is of particular interest as numerous studies reported the role of IL-1β in neuroinflammation and neurodegeneration especially within the field of Alzheimer’s disease: for example, postmortem brain sample analysis revealed elevated levels of Il-1β production, including around amyloid beta plaques; all suggestive of pathology-associated inflammation (33–35). Hence, we speculate that in CSF1R+/- mutant microglia, which already have a predisposition towards increased secretion of IL-1β, are constantly being activated by IL-1β, either autocrine or paracrine, creating a dysfunctional neuro-environment that contributes to HDLS pathology. As such, experimental models looking to block or reduce IL-1β levels in the brain using the human mAb, canakinumab, targeting the proinflammatory cytokine IL-1β (36, 37), during both health and disease could be a potential target for therapy and may help to ameliorate HDLS pathology.

While no model perfectly recapitulates the full complexity of a disease, the application of iPSC-derived organoid co-culture approaches to primary patient-derived cells offers a valuable way to study the disease mechanisms of this rare pathology using human cells. Future studies employing microglia-sufficient patient-derived brain organoids have the potential to shed much-needed light into the pathogenesis of HDLS and other neurodegenerative diseases.

eLife Assessment

This important study sheds light on the mechanisms underlying a rare brain disease, offering insight into the role of microglia in this complex pathophysiology. The evidence presented is solid, utilizing state-of-the-art laboratory models to explore cellular interactions and disease development. While further research is needed, this study will be of interest to neuroscientists and clinicians aiming to understand and combat similar neurodegenerative disorders.

Reviewer #1 (Public Review):

Here, using an organoid system, Wong et al generated a new model of hereditary diffuse leukoencephalopathy with axonal spheroids, with which they investigated how CSF1R-mutaions affect the phenotypes of microglia/macrophages, and revealed metabolic changes in microglia/macrophages associated with a proinflammatory phenotype.

In general, this paper is interesting and well-written, and tackles important issues to be addressed.

This study suffers from several major concerns and limitations that dampen the value of the study. As the authors also mentioned, models that perfectly recapitulate the complexity of the HDLS brain the models would be required to better understand the molecular mechanisms of the disease. In this regard, it is unclear how nicely the organoid system in this study can recapitulate the condition in patients with HDLS (e.g. reduced microglia density, downregulated expression of P2YR12, pathological alterations). In addition, the authors used two different models with distinct mutations that could produce different readouts in CSF1R-mediated cellular responses.

Although the reviewer does understand the importance of providing several options/tools to study rare diseases like HDLS and the difficulty of generating stable organoids with less variation, it is unclear if the different outcomes between HD1 and HD2 are generated through different mutations or simply due to different differentiation efficiency from iMacs (e.g. Figure 2B), which needs to be confirmed. Lastly, there is an over-interpretation regarding the results in Figure 6A. There is no difference between isoHD1 iMac control and HD1 Mut iMac.

Reviewer #2 (Public Review):

Summary:

This paper investigates a rare and severe brain disease called Hereditary Diffuse Leukoencephalopathy with Axonal Spheroids (HDLS). The authors aimed to understand how mutations in the gene CSF-1R affect microglia, the resident immune cells in the brain, and which alterations and factors lead to the specific pathophysiology. To model the human brain with the pathophysiology of HDLS, they used the human-specific model system of induced pluripotent stem cell (iPSC)-derived forebrain organoids with integrated iPSC-derived microglia (iMicro) from patients with the HDLS-causing mutation and an isogenic cell line with the corrected genome. They found that iPSC-derived macrophages (iMac) with HDLS mutations showed changes in their response, including increased inflammation and altered metabolism. Additionally, they studied these iMacs in forebrain organoids, where they differentiate into iMicro, and showed transcriptional differences in isolated iMicro when carrying the HDLS mutation. In addition, the authors described the influence of the mutation within iMicro on the transcriptional level of neurons and neural progenitor cells (NPCs) in the organoid. They observed that the one mutation showed implications for impaired development of neurons, possibly contributing to the progression of the disease.

Overall, this study provides valuable insights into the mechanisms underlying HDLS and emphasizes the importance of studying diseases like these with a suitable model system. These findings, while promising, represent only an initial step towards understanding HDLS and similar neurodegenerative diseases, and thus, their direct translation into new treatment options remains uncertain.

Strengths:

The strength of the work lies in the successful reprogramming of two HDLS patient-derived induced pluripotent stem cells (iPSCs) with different mutations, which is crucial for the study of HDLS using human forebrain organoid models. The use of corrected isogenic iPSC lines as controls increases the validity of the mutation-specific observations. In addition, the model effectively mimics HDLS, particularly concerning deficits in the frontal lobe, mirroring observations in the human brain. Obtaining iPSCs from patients with different CSF1R mutations is particularly valuable given the limitations of rodent and zebrafish models when studying adult-onset neurodegenerative diseases. The study also highlights significant metabolic changes associated with the CSF1R mutation, particularly in the HD2 mutant line, which is confirmed by the HD1 line. In addition, the work shows transcriptional upregulation of the proinflammatory cytokine IL-1beta in cells carrying the mutation, particularly when they phagocytose apoptotic cells, providing further insight into disease mechanisms.

Weaknesses:

The authors have not elucidated the significance of the increased CSF1 dosage in Figure 2F, aside from its effect on cell viability, lacking a thorough discussion of this result. Additionally, while transcriptomic and metabolic alterations related to the mutation were demonstrated in iMac models, similar investigations in iMicros are absent, necessitating further experiments to validate the findings across cell models. The conclusion drawn regarding cytokine levels lacks robust support from the data, particularly considering the varied responses observed in different mutant lines. Further analysis of the secretome (e.g. via ELISA) could provide additional insights. Moreover, the characterization of iMicros is incomplete, with limited protein-level analysis (e.g. validate RNA-seq via flow cytometry). Additionally, the claim of microglial-like morphology lacks adequate evidence, as the provided image is insufficient for such an assessment. RNA-seq experiments should be represented better, it is not possible to read the legends or gene names in the figures. Maybe the data sets can be combined into PCAone and one overall analysis, e.g. via WGCNA-like analyses? This would make it easier for the reader to compare the two cell lines side by side. Furthermore, inaccuracies and omissions in the figure legends compromise the clarity of data representation. Statistical test information is missing. Finally, inconsistent terminology usage throughout the paper may confuse readers (iMac versus iMicros).

Author response:

The following is the authors’ response to the original reviews.

We thank the reviewers for their feedback on our manuscript. Taking the advice of the reviewers, we have streamlined the text and formatted the figures to conform to the format instructions. We believe that the revised manuscript has been improved.

Point-by-point responses are presented below.

Reviewer #1:

(1) There is an over-interpretation regarding the results in Figure 6A. There is no difference between isoHD1 iMac control and HD1 Mut iMac.

We thank the reviewer for his/her feedback on our manuscript. We have since changed the wordings on Page 11, line 294 of the manuscript, to reflect this important point.

Reviewer #2:

(2) The authors have not elucidated the significance of the increased CSF1 dosage in Figure 2F, aside from its effect on cell viability, lacking a thorough discussion of this result.

We have incorporated the significance of the results of our CSF1 dosage data with a newly added observation of an upregulated immature myeloid marker and downregulated expression mature macrophage marker within mutant iMac from the respective RNA-seq data (Page 5, line 163); and elaborated further within the Discussion section that this results in the possible generation of immature iMacs even after maturation (Page 14, line 356).

(3) Additionally, while transcriptomic and metabolic alterations related to the mutation were demonstrated in iMac models, similar investigations in iMicros are absent, necessitating further experiments to validate the findings across cell models.

We thank the reviewer for this feedback and feel that this is beyond the scope of this study at current stage, and that we would keep this in consideration to incorporate into subsequent experiments.

(4) The conclusion drawn regarding cytokine levels lacks robust support from the data, particularly considering the varied responses observed in different mutant lines. Further analysis of the secretome (e.g. via ELISA) could provide additional insights.

We thank the reviewer for this feedback and feel that this is beyond the scope of this study at current stage, and that we would keep this in consideration to incorporate into subsequent experiments.

(5) Moreover, the characterization of iMicros is incomplete, with limited protein-level analysis (e.g. validate RNA-seq via flow cytometry).

We thank the reviewer for this feedback and feel that this is beyond the scope of this study at current stage, and that we would keep this in consideration to incorporate into subsequent experiments.

(6) Additionally, the claim of microglial-like morphology lacks adequate evidence, as the provided image is insufficient for such an assessment.

We have added confocal images depicting microglial-like morphology in our co-culture system within Supp Fig 3C.

(7) RNA-seq experiments should be represented better, it is not possible to read the legends or gene names in the figures. Maybe the data sets can be combined into PCAone and one overall analysis, e.g. via WGCNA-like analyses? This would make it easier for the reader to compare the two cell lines side by side.

We have since enhanced the quality of the respective RNA-seq figures with enlarged data points and gene names for better clarity.

(8) Statistical test information is missing.

We are sorry for leaving this out and have added the statistical test information within Page 15 of the methods section.

(9) Finally, inconsistent terminology usage throughout the paper may confuse readers (iMac versus iMicros).

We have streamlined the terminology used within Page 10, line 265 and 267, of the manuscript for better consistency.

(10) Fig. 1D: which cell line is displayed here?

Mut HD1 iPSC is displayed here. We have also revised the figure legend of Fig 1D within Page 1, line 8 to include this information.

(11) Fig. 1E: Karyotype of which cell line is shown?

We have included karyotype of both IsoHD1 and IsoHD2 iPSC in Fig 1E, and also revised the legend within Page 1, line 11, to reflect this change.

(12) Supp. Fig. 1: scale bar information missing.

We thank the reviewer for pointing out this and have revised the legend within Page 1, line 17, to include scale bar information.

(13) Fig. 5: legend for A is missing.

We thank the reviewer for pointing out this and have revised the legend within Page 2, line 91, to include Figure (A) within.

1. Supp. Fig. 3A says 30 days, but only 23 days are shown.

We are sorry for making this inadvertent typo and have since aligned the correct days (31 days) shown within the figure (Supp Fig 3A) and legend (Page 3, line 110, 113), as mentioned in the manuscript.

(15) Supp. Fig. 3C: scale bar length is incorrect.

We did a recheck and are confident that the scale car is of the correct length. The images displaying the respective fluorescent channels are proportionately reduced with respect to the main figure (now Supp. Fig. 3D), and thus are of the same size (200 uM).

(16) Fig. 6: legend for D, E is missing.

We have revised the figure legend within Page 3, line 128, 130 and 131, to address said missing legends.

(17) Stem cells do also express Sox2. how does Sox2 expression lead to the conclusion of an optimal generated organoid?

We thank the reviewer for pointing this out. Sox2 has been defined as a core intrinsic factor for regulating pluripotency (Avilian et al, 2003, Zhang et al, 2014), as well as lineage specifiers to regulate ectodermal differentiation which is crucial in controlling neural initiation and differentiation from iPSC (Zhao et al, 2004, Thomson et al, 2011, Wang et al 2014). Additionally, Sox2 is highly expressed in proliferating neural progenitor cells as documented in previous iterations of cerebral organoids generation protocol (Lancaster et al 2013, Qian at el, 2018). Perhaps “optimally” sounds too forced in this context, as such we have toned down on the phrasing.

(18) HD1 and HD2 react differently (e.g. in IL-1B production), but the text is written often as if both cell lines react in the same way.

We thank the reviewer for pointing this out and have since clarified this within Page 4, line 366-368, of the manuscript.

(19) Precise information on medium missing (e.g. no Pen/Strep?).

We thank the reviewer for pointing out this. Culturing of iPSC colonies was done without the use of Pen/Strep. Additionally, we have elaborated the medium composition for our iMac cultures for clarity within Page 4, line 106, of Materials and Methods as well as the information within Supp. Table 4.

(20) How was ReleSR used exactly?

We have included the usage of ReleSR within Page 2, line 41 of Materials and Methods.

(21) What kind of microscopes/objectives were used for imaging?

We have added the respective microscope details for bright-field, phase-contrast and cytospin related experiments within Page 3, line 73, and Page 14, line 360, of Materials and Methods.

(22) For the dissociation of organoids: what kind of pipit was use and at which temperature were organoids incubated?

We have included the pipette used for organoids dissociation, as well as the incubation temperature for organoids culture within Page 9, line 243, 244 and 245, of Materials and Methods.

(23) How was the RNA-seq analysis done? Which packages? Which versions?

We provide now the information requested in the material and method section.

Version 2

Abstract

Abstract

Hereditary diffuse leukoencephalopathy with axonal spheroids (HDLS) is a rare, fatal, adult-onset neurodegenerative disease that is most often caused by mutations affecting the Colony Stimulating factor-1 Receptor (CSF-1R). To understand how CSF-1R-mutation affects human microglia – the specialized brain-resident macrophages of the central nervous system – and the downstream consequences for neuronal cells, we used a macrophage and forebrain organoid co-culture system based on induced pluripotent stem cells generated from two patients with HDLS, with CSF-1R gene-corrected isogenic organoids as controls. Macrophages derived from iPSC (iMacs) of patients exhibited a metabolic shift towards the glycolytic pathway and reduced CSF-1 sensitivity, which was associated with higher levels of IL-1β production and an activated inflammatory phenotype. Single-cell RNA sequencing revealed that iMacs adopt a reactive state that leads to impaired regulation of neuronal cell populations in organoid cultures, thereby identifying microglial dysregulation and specifically IL-1β production as key contributors to the degenerative neuro-environment in HDLS.

Introduction

Hereditary diffused leukoencephalopathy with axonal spheroids (HDLS) – also known as adult-onset leukodystrophy with axonal spheroids and pigmented glia (ALSP), is a rare, inherited, autosomal dominant, neurodegenerative disease that is characterized by patchy axonal swellings (spheroids) and demyelination that result in the alteration of the white matter of the brain (1). Although detailed epidemiological data are lacking, in 2021, it was estimated that approximately a quarter of a million people globally are affected by the condition (2); yet our understanding of the pathogenesis of the disease is limited, and there is no cure or even effective treatment.

For more than a decade, we have known that HDLS is caused by mutations affecting the colony-stimulating factor-1 receptor (CSF-1R) (3) which most often occur in the tyrosine kinase domain (TKD) (4), and lead to reduced auto-phosphorylation in response to CSF-1, which results in impaired downstream signaling (4, 5). These mutations most profoundly affect microglia (6), the specialized brain-resident macrophages that play critical roles in brain development and homeostasis (7, 8) as well as in a wide range of neuroinflammatory and neurodegenerative diseases (9). Patients with HDLS have few, small, microglia that are distributed abnormally within the neocortex and that express low levels of the homeostatic marker P2RY12 and high levels of the inflammatory marker CD68, compared to healthy controls (10–12).

Studies aiming to understand the pathophysiology of HDLS have used various approaches. Rodent models have been generated, but do not faithfully recapitulate the disease phenotype in humans: the earliest haplo-insufficient Csf-1r+/- mice model exhibited microgliosis within brain regions (13), but this was not recapitulated in a later Csf-1r knockout rat model (14); while a subsequent orthologous mouse model has phenocopied the reduced microglia density seen in patients with HDLS but does not exhibit brain pathology (15). A zebrafish model of HDLS similarly saw reduced microglia during brain development (16), but does not offer the opportunity to study pathology associated with adult-onset neurodegenerative diseases (17). To overcome these species-specific limitations, in this study we generated human induced pluripotent stem cell (iPSC) - derived macrophages from patients with HDLS and used them to create microglia-sufficient autologous brain organoids (18) in which we characterized the effects of naturally occurring CSF-1R mutation on microglial function. By comparing iPSC-derived macrophages to genetically corrected macrophages from the same patients, we uncovered evidence of significant transcriptional and metabolic reprogramming that was associated with profound changes to neuronal regulation by these cells, indicative of the likely mechanistic links between microglial dysfunction and HDLS.

Results

Generation and characterization of iPSC lines from patients with HDLS

We first isolated dermal fibroblasts from skin biopsies taken from two patients with HDLS and confirmed mutations in the TDK region of the CSF-1R gene (HD1, HD2; clinical characteristics shown in Supplemental Table 1) then cultured them for approximately four weeks before reprogramming into induced pluripotent stem cells (iPSCs) (Fig 1A). Four to five weeks post-transfection, we observed iPSC-like colonies emerging (Supp Fig 1B), and 7-10 days later, we selected those clones that were proliferating and displayed typical iPSC morphology (Fig 1C). We confirmed successful reprogramming by visualization of pluripotency markers including SOX2 and OCT4 (nuclear markers), and TRA-1-60 and SSEA-4 (intracellular marker), by immunofluorescence (Fig 1D). We also confirmed that these clones had normal karyotype (Fig 1E) and termed them Mut HD1 and Mut HD2. One clone per cell line was used for further analysis. We then generated isogenic from Mut HD1 and Mut HD2 iPSC cell lines which we termed IsoHD1 and IsoHD2 respectively, as controls (see Methods). Clones that were karyotypically normal were used for downstream experiments alongside their mutant counterparts (Fig 1B), with no off-target events identified (Supp Fig 1C). Collectively, HDLS iPSC and their corresponding isogenic iPSC controls were successfully generated from two patients with HDLS (HD1 and HD2).

Generation and characterization of iPSC lines from patients with HDLS

A) Schematic representation of the generation of patient-derived induced pluripotent stem cells (iPSC) from dermal fibroblasts, and the subsequent generation of isogenic controls via gene editing with CRISPR/Cas9. B) Sequencing of the genomic CSF-1R locus in donor-derived iPSC displaying mutant allele or genetically edited allele in isogenic control cell lines. C) Phase-contrast microscopy of mature reprogrammed iPSC colonies displaying typical hESC morphology. Images were acquired with a standard microscope (Nikon) with a 10 × objective. Scale bar represents 250 µM. D) Immunofluorescence imaging of Mut HD1 iPSC colonies for the pluripotency markers SOX2, OCT4, TRA-1-60 and SSEA-4. Scale bar represents 150 µM and 100 µM for left and right panel respectively. E) Cytogenetic analysis of IsoHD1 (left) and IsoHD2 (right) iPSC clone showing a normal karyotype.

Generation and characterization of patient iPSC-derived macrophages

We next sought to differentiate primitive microglia-like macrophages from patient-derived iPSC cell lines (iMacs) using our serum-free protocol that faithfully recapitulates the generation of in vivo microglial progenitor cells (19). Briefly, iPSCs were initially inducted towards the mesodermal lineage before being differentiated into hemangioblast-like cells with the potential to differentiate into both endothelial and hematopoietic progenitors - the latter emerging as free-floating cells as early as day six after seeding - which we collected and then terminally differentiated into iMacs by the addition of CSF-1 (Fig 2A). Although HDLS mutation downregulates the sensitivity of CSF-1R to its ligand (5), we were able to successfully generate iMacs using this protocol: the yield of iMacs from iPSC varied between cultures (Fig 2B), but flow cytometry confirmed their consistent macrophage-like phenotype, with abundant expression of CD45 and CD14 (Fig 2C), and Giemsa staining showed that iMacs exhibited comparable morphology between donors (Fig 2D). Functionally, all iMac cultures were equally able to take up PE-labelled beads that were detectable within their cell bodies (Fig 2D); flow cytometry confirmed that the level of actin-dependent phagocytosis of pHrodo beads was comparable among CD45-expressing cells from all cultures (Fig 2E). Collectively, these data show that iMacs generated from HDLS donor-derived iPSC adopted the typical morphology, phenotype and functions associated with macrophages.

Generation and characterization of patient iPSC-derived macrophages

A) Schematic representation of the derivation of primitive macrophages from donor-derived iPSC. B) iMacrophage (iMac) yields across all cell line variants. C) Flow cytometry analysis of iPSC and iMac expression of macrophage surface markers CD45 and CD14. D) Morphology of iMacs visualized by Giemsa staining (left panel), and confocal imaging (middle panel); and internalization of PE-conjugated beads visualized by confocal imaging. Scale bar represents 30 µM. E) Uptake of pHrodo-labeled beads by iMacs in the presence and absence of cytochalasin D (Cyto D). F) Cell viability of iMacs used to determine CSF-1 sensitivity via an MTT assay, after incubating with varying concentrations of CSF-1 for seven days. Data are presented as mean ±SEM.

The mutations associated with HDLS downregulate the sensitivity of CSF-1R to its ligand CSF-1, but to different extents depending on the mutation site (5). We therefore compared the survival of our mutant and isogenic cells lines from both patients during culture with CSF-1. As expected, even with high concentrations of CSF-1, relatively few mutant cells survived across seven days of culture; by contrast, even low CSF-1 levels allowed significantly more isogenic cells to survive (Fig 2F).

In summary, we generated iMacs from iPSC of patients with HDLS and their respective isogenic controls that exhibit typical macrophage morphology, phenotype, and phagocytic capabilities.

Mutant and isogenic control iMacs have distinct gene expression profiles

Although iMacs derived from patients with HDLS exhibited core macrophage features, we wanted to more fully understand the impact of CSF-1R mutation on these cells. We first characterized their gene expression profile using bulk RNA-seq and compared it to that of their respective CSF-1R-restored isogenic iMac cultures.

Principal component analysis (PCA) showed that patterns of gene expression differed markedly between mutant and isogenic iMacs. Among cell cultures from HD1, mutant iMacs exhibited patterns of gene expression consistent with a more reactive state compared to isogenic CSF-1R-restored HD1: HD1 Mut had higher expression of genes involved in pathways related to proliferation (IGF2, MYBL2, STAT1, INCEP, TGFA) and the defense response to virus (IFI27, IFI44L, STAT1, MX1, IFIT3); while Iso HD1 iMacs expressed high levels of genes involved in chemokine-related responses and signaling pathways (CXCL2, ADAM8, CCL31L, CCL5,), as well as leukocyte migration and chemotaxis (CXCL3, CCL13, CCL2, E5AR1, C3AR1) (Fig 3A-C). We observed a similar transcriptomic profile in HD2-derived iMacs: HD2 Mut showed up-regulated gene expression for the defensive response to virus, while isogenic HD2 iMacs exhibited high expression levels of genes involved in leukocyte migration and chemotaxis (Fig 3D-F). These data are in line with previous reports suggesting that macrophages from patients with HDLS adopt an inflammatory phenotype with increased expression of the immunoreactive markers CD68 and CD163 (10, 12). Additionally, an upregulation of the immature myeloid cell marker MPO (20), and a reduced expression of mature macrophage marker MARCO (21) was observed within mutant iMac of both cell lines. Taken together, iMac cultures derived from patients with HDLS exhibited transcriptomic profiles consistent with upregulated inflammatory gene expression as a result of impeded CSF-1R function.

Mutant and isogenic control iMacs have distinct gene expression profiles

A and D) PCA analysis, B and E) volcano plots and C and F) GO analysis, of mutant and isogenic iMacs derived from HD1 and HD2 differentially expressed genes (DEGs) respectively. Representative curve of ECAR (G, O) and their respective quantified data (H-J, P-R) are shown. Representative curve of OCR (K, S) and their respective quantified data (L-N, T-V) are shown. Real-time assessment of bioenergetic profile between mutant and isogenic HD1 iMacs, measured using extracellular flux assay and a mitochondrial stress test. The following glycolytic parameters were calculated based on ECAR: B) glycolysis, C) glycolytic capacity, D) glycolytic reserve. Data shown as means + SEM, n=9. The following parameters were calculated based on OCR: F) basal, G) ATP production and H) maximal respiration. 2-DG = 2-deoxy-glucose, FCCP = carbonyl cyanide 4-trifluoromethoxy-phenylhydrazone. (n=3) Data are presented as mean ± SEM.

Mutant and isogenic control iMacs have distinct metabolic profiles

Studies have shown that activated macrophages often undergo metabolic reprogramming to fulfil the energetic needs of demanding biochemical processes (22, 23); alongside, we know that energy metabolism, as well as metabolic cell features, plays an important role in mediating macrophage function and plasticity (24). Therefore, we next compared the cellular metabolic profiles of patient-derived CSF-1R mutant and isogenic CSF-1R-restored iMacs.

We first assessed glycolysis by measuring the extracellular acidification rate (ECAR) in the presence of specific modulators (glucose, oligomycin, 2-deoxy-glucose) that were added sequentially to the cell cultures to reveal various aspects of the metabolic pathway (see Methods): this showed that CSF-1R-mutant iMacs from both HD1 and HD2 had a significant overall increase in glycolytic function (Fig 3G and 3O). Although the magnitude of the difference between iMac HD1 and HD2 and their respective CSF-1R-restored counterparts varied, in both cases, CSF-1R mutation was associated with significantly increased glycolytic capacity and glycolytic reserve.

Next, we determined the functional metabolic profile of mitochondrial respiration by measuring real-time changes in the extracellular oxygen consumption rate (OCR) during sequential treatment of cells with oligomycin, carbonyl cynide p-trifluoromethoxyphenylhydrazone (FCCP) and a combination of rotenone and antimycin A. As before, although the magnitude of effect was less for cells from HD1, both sets of iMacs showed a similar trend towards higher OCR values, and greater capacity for basal respiration, ATP production, and maximal respiratory capacity in the presence of mutated CSF-1R in the case of HD2 Mut cells, these differences achieved statistical significance across all three bioenergetic parameters.

Taken together, these results indicate that the absence of fully functional CSF-1R in cells from HD1 and HD2 is associated with an upregulated glycolytic metabolic profile, which is cells from HD2 is clearly accompanied by a higher mitochondrial oxidative phosphorylation rate. This suggests that macrophages from patients with HDLS have an altered metabolic profile that correlates with their higher activation status and could contribute to disease pathology.

HDLS iMacs upregulate IL-1β when exposed to apoptotic neuronal cells

Our data suggest that mutant iMacs have dysregulated transcriptomic and metabolic profiles, suggestive of an active inflammatory state. To better understand how this would be expressed in the context of their functions in the brain, we mimicked a need for apoptotic cell clearance by incubating iMacs from patients with HDLS with UV-irradiated neuronal SH-SY5Y cells (Supp Fig 2A). Both iMac lines (CSF-1R mutant and isogenic) from both patients effectively phagocytosed apoptotic but not control, non-UV-irradiated, SH-SY5Y cells, in an actin-dependent manner (Supp Fig 2B).

We then measured the expression of genes encoding pro-(IFN-g, TNF-α, IL-1β, IL-6, IL-18) and anti-(IL-10, IL-12, TGF-β) inflammatory cytokines following incubation of iMacs with UV-irradiated or non-irradiated SH-SY5Y cells, or without any SH-SY5Y cells, for eight hours. We observed a significant and specific increase in IL-1β transcription in CSF-1R mutated iMacs from both HD1 and HD2 when exposed to apoptotic cells, compared with their isogenic counterparts. Among the anti-inflammatory cytokines, Mut HD2 cells exhibited significantly increased transcription of the gene encoding TGF-β when cocultured with UV-treated cells, alongside high baseline levels of IL-12 and TGF-β in control cultures compared to isogenic HD2 iMacs. Interestingly, all CSF-1R mutant iMac groups had higher levels of transcription of most pro-inflammatory cytokines at baseline, consistent with their activated/reactive transcriptional profile, than did their CSF-1R-restored counterparts (Fig 4A-B).

Cytokine profiling of iMacs saw upregulated levels of IL-1β when probed with apoptotic neuronal cells

(A-B) Levels of transcription of pro- and anti-inflammatory cytokine genes (IFNg, TNFa, IL-1b, IL-6, IL-18, IL-10, IL-12 and TGFb) quantified by qPCR. (+uv depicts UV-treated, -uv depicts non-UV treated). n=3 per cell line. Data are presented as mean ±SEM.

Collectively, all iMacs exhibited the ability to specifically phagocytose apoptotic cells, but in the absence of a fully functional CSF-1R, this led to high levels of IL-1β transcription, which reflected a general propensity towards the transcription of pro-inflammatory cytokine genes, even in the absence of overt stimulation. Importantly, restoration of CSF-1R via gene editing was sufficient to reverse this inflammatory phenotype.

iMacs differentiate into microglia-like cells (iMicro) in a coculture with forebrain organoid

During HDLS, patients typically exhibit degeneration of the frontal lobes of the brain, which contributes to the deficit and decline in cognitive ability, changes to personality and behavior, and dysfunctional motor, social, and language skills (25, 26). Therefore, we next generated iPSC-derived forebrain organoids including autologous iMacs from patients with HDLS to better understand their potential role in the brain.

To better understand how diseased or non-diseased macrophages affect the developing brain, we first generated forebrain organoids from the isogenic cell line variant of the respective donor’s iPSC (see Methods). After 31 days, organoids were harvested (Supp Fig 3A) and co-cultured for seven days with isogenic or mutant iMacs that had been generated in parallel (Fig 5A). At this point, each organoid was removed from its coculture and cultured individually for 16 more days (Fig 5B).

iMacs differentiate into microglia-like cells in coculture with forebrain organoids

A) Schematic diagram of co-culture between Day 31 iMacs and forebrain organoid. B) Forebrain organoids depicted before, during and after co-culture with iMacs. Day 31 isogenic forebrain organoids derived from the respective donor iPSC cell line were cocultured with the same isogenic- and mutant-cell-line-derived Day 31 iMacs. 30,000 iMacs were cocultured with each respective organoid derived from HD1 and HD2. Figure shows matured organoids before and after the addition of iMacs during coculture. C) Three-dimensional visualization of co-cultured brain organoids after 23 days. Neurons were labeled for NESTIN and NPCs for SOX2. Microglia-like cells (IBA-1+) were visualized on the surface of organoids. D) Flow cytometry analysis of neurons, NPCs, and macrophage populations within forebrain organoids after 23 days of coculture. E-F) Donor-derived iMacs in forebrain co-cultures undergo distinct changes in gene expression as observed in PCA analysis (E and H), volcano plots (F and I) and GO analysis (G and J) in both HD1 and HD2 variants. Scale bar represents 200 µM.

We then applied 3D imaging of forebrain organoids to assess their morphology and the distribution/differentiation of the iMacs. This showed that iMacs expressing the microglial marker IBA-1, were mostly located on the surface of the organoids (Fig 5C). We also labelled organoids for NESTIN (a neuronal cell marker) and SOX2 (expressed by neuronal progenitor cells (NPCs). In all of the organoids, neural epithelial rosettes were present, consisting mostly of SOX2+ NPCs (Supp Fig 3B); these structures indicate favorable organoid growth as they demonstrate proper lineage progression leading to forebrain region identity (27) Importantly, organoids to which we did not add iMacs did not contain any IBA-1+ cells (Supp Fig 3C); this was as expected from our previous work (18), but was in contrast to previous studies that reported innately-derived microglia in mesodermal-lineage-containing cerebral organoids (27, 28).

Next, we characterized the different cell populations within organoids after 23 days of coculture and compared iMacs before and after co-culture. After organoid dissociation, cells were labelled with antibodies to differentiate neurons (CD45-/CD184-/CD44-/CD15lo/CD24+), NPCs (CD45-/CD184+/CD271-/CD44-/CD24+) and iMacs (CD45+) before undergoing sorting (Fig 5D) followed by bulk-RNA sequencing. During the course of the work, sufficient iMicro numbers for downstream analysis were only generated from coculture with mutant iMacs for HD1 and isogenic iMacs for HD2 respectively. Compared to Day 31 non-co-cultured iMacs, HD1 iMacs after co-culture had profoundly altered their gene expression (Fig 5E), with the upregulated DEGs showing an increase in various pathways that are characteristic of microglial functions in the brain such as central nervous system and brain development (CNTN1, ZIC2, EFNA2, GLI3), synapse organization, synapse and neuron organization and development (SLC1A1, DGKB, TRIM67), as well as neurogenesis (CHD11, ADCYAP1) (Fig 5F-G).

Likewise, for HD2 iMacs, PCA also suggested that their patterns of gene expression were markedly different between cocultured and non-cocultured populations (Fig 5H), while GO analysis of the upregulated DEGs revealed an increase in various pathways linked with typical microglial functions in the brain such as glial cell differentiation and gliogenesis (CLU, MDK, CXCR4, RELN, TMEM98, TUBA1A), axonal development (UCHL1, MAP1B, CRABP2, TUBB2B, LGI1) as well as forebrain development (TUBB2B, MDK, CXCR4, RELN, LHX2) (Fig 5I-J). This shows that iMacs undergoing coculture with forebrain organoids have differentiated to a certain extent, into microglia-like cells.

Co-culture results in impaired regulation of neurons within organoids

Microglia not only act as sentinel cells, but are also important contributors to neuroplasticity and neurogenesis in the developing brain that ultimately help shape the brain circuitry (29). Part of this role requires microglia to regulate the population of NPCs and neurons during the unrestrained, highly proliferative phase of early brain development (30, 31); therefore, we next asked whether the microglia-like cells from our iMac-forebrain organoids were also capable of regulating neurogenesis in vitro.

As such, we sought to compare the size of the organoids, as well as the individual major cell population in each organoid, for both variants between isogenic organoids that did not undergo co-culture, and those that underwent 23 days of co-culture by comparing their area before and after co-culture. Co-culture of HD1 saw a slight reduction in the organoid size among both co-culture parameters before and after co-culture. However, comparison of the organoid size on coculture Day 23 only, saw the isogenic control parameter as the only group that had a reduction in organoid size when compared to the organoid-only control (Fig 6A). On the contrary, HD2 displayed no significance in organoid size reduction even after co-culture (Fig 6B).

Co-culture results in impaired regulation of neurons within organoids

Size of A) HD1 and B) HD2 organoids measured before and after co-culture with mutant and isogenic iMacs in both HD1 and HD2 variants. The effects of co-culture on neuronal populations in C) HD1 and D) HD2 organoids after 23 days. Transcriptomic analysis of co-cultured E) neurons and F) NPCs from HD2 organoids after 23 days. Data are presented as mean ±SEM.

Subsequently, these co-cultured organoids were dissociated, stained, sorted and quantified for the respective cell populations. Co-culture of IsoHD1 organoid with IsoHD1 iMac saw a significant reduction in the neuronal population when compared to control organoids that did not underwent co-culture. Conversely, coculture with mutant HD1 iMac saw a smaller degree of neurons reduction but was not significant (Fig 6C). Next, co-culture of IsoHD2 organoid with IsoHD2 iMac saw a slight reduction in neuronal population that was not statistically significant, while the co-culture with mutant HD2 iMac saw no difference when compared with control organoids (Fig 6D). Interestingly, we did not detect any change in the NPC population for either HD1 or HD2 after co-culture, likely because their numbers were relatively low to begin with due to the cellular composition of the forebrain organoid. Similarly, the iMac population was also small in all co-cultures. Transcriptomic analysis of the neurons of HD2 co-culture saw the mutant iMac co-culture having upregulated expression of genes associated with mitochondrial respiration (BCS1L, MRPL18, TSFM, MRPL11, MRPS30) but this phenomenon was not present in the isogenic iMac co-culture (Fig 6E). Conversely, transcriptomic analysis of the NPCs of HD2 co-culture saw the isogenic iMac co-culture having upregulated expression of genes associated with mitochondrial respiration (Fig 6F). Taken together, our model suggests that HDLS mutation affects neuronal population in a microglia-dependent manner and that this phenomenon is variable and mutation-dependent.

Discussion

HDLS is an incurable neurodegenerative disease that begins in adulthood, progresses rapidly and has a grim prognosis. Efforts to elucidate the functional or dysfunctional aspects of the causative mutations in the CSF-1R gene have been hampered by the lack of an accurate disease model that recapitulates the human pathology: although rodent and zebrafish models have been generated and imparted valuable insights towards HDLS (13–16), discrepancies and limitations in each case have led to incomplete recapitulation of the disease’s pathology, specifically in microglia numbers within brain regions and the effects on respective cognitive and sensorimotor tests. Additionally, a model looking into immune-neuronal crosstalk to better understand a neurodegenerative disorder that is primarily associated with microgliopathy, is important to advance the field. Here, we first generated iPSC lines from two patients with HDLS and then used CRISPR/Cas9 technology to reinstate a fully functional CSF-1R gene into some of the lines. While the generation of an HDLS donor-derived iPSC cell line has been reported (32), no other in vitro studies have yet reverted the causative mutation to wild type; a necessary step to understanding the direct impacts of diminished/absent CSF-1R signaling.

We then generated and characterized iMacs - precursors of microglia - from these lines, which exhibited typical morphology, phenotype and phagocytic capacity that were not obviously affected by CSF-1R mutation. Although we were able to generate phenotypically and functionally similar iMac across these lines for example for phagocytic capability, iMac yield varied greatly between the cell lines with a 50% lower yield for Mut HD1 compared to IsoHD1 and, conversely, a higher yield for Mut HD2 200% compared to IsoHD2. This fundamental difference could be attributed to their respective mutations: the nonsense mutation identified in HD2 could not only more profoundly affect receptor signaling, but might also be a major factor during the course of gene editing leading to incomplete reversion to wild type at the epigenetic level, which we speculate was not the case for IsoHD1. Additionally, this observation could also be attributable to a reduced CSF-1 sensitivity in HD2 Mut iMacs as compared to HD1 Mut iMacs and we speculate that this would potentially give rise to the generation and subsequent maintenance of immature iMacs that lacks maturation even after terminally differentiated into tissue-resident macrophages.

Functionally, while phagocytosis of E.Coli beads was similar across lines, the baseline cytokine production as well as after phagocytosis of apoptotic neuronal cell line was different with an increase of IL-1β secretion in both mutant HD1 and HD2, compared to isogenic HD1 and HD2 iMacs. In the case of phagocytosis of apoptotic cells, this was not due to a difference in phagocytic activity as all cells were able to phagocytose apoptotic cells at a similar level. Importantly, higher levels of baseline pro-and anti-inflammatory cytokine production in the control group without apoptotic cells were observed within HD1 mutant, and both HD1 and HD2 mutant iMacs respectively. This suggest that mutant iMacs are innately in a more reactive or activated state when compared to isogenic iMacs and could reflect the reactive phenotype of HDLS-associated microglia and macrophage identified in vivo (10, 12). Conversely, isogenic iMacs seems to adopt less of the reactive phenotype with the reverted mutation as identified with an overall lesser cytokine production level. Importantly, such reactive phenotype was also highlighted by our transcriptomic and metabolic profiling with upregulated pathways involving pro-inflammatory cytokines and preferential utilization of the glycolytic pathways in the mutant groups, a feature of macrophage activation (33, 34), respectively.

We subsequently incorporated donor-specific and their isogenic controls into a co-culture system using 3D forebrain organoids and iPSC-derived macrophages to better study HDLS. Isogenic and mutant iMacs co-cultured separately with their respective isogenic forebrain organoid differentiate into microglia-like cells with 23 days of coculture as shown by the expression of the microglial marker IBA-1, their ramified morphology as well as their transcriptomic profiles with up-regulation of genes associated with biological pathways that are typical of microglia functions such as neurodevelopment, neurodifferentiation, neurogenesis and brain development.

Subsequently, we analyzed the size of the organoids before and after co-culture as we recently showed that addition of iMac modulated NPC differentiation, limiting their proliferation and promoting axonogenesis, resulting in a reduction of co-culture organoid size (18). Here, while we noticed a significant decreased in the size of isogenic HD1 forebrains co-cultured with isogenic HD1 iMac compared to isogenic HD1 organoids, we did not measure any significant change in organoid size in the isogenic HD1 organoids co-cultured with mutant HD1 iMacs. This suggested that the crosstalk between iMicro and NPC is dysregulated in the context of the HD1 mutation. Since we showed that such size reduction is mediated by cholesterol transfer from iMicro to NPC, further work will be required to characterize cholesterol metabolism of HD iMac.

While we did not observe major changes in cell type composition in these cocultures, transcriptomic analysis of neuronal and NPC population after co-culture showed an increase in mitochondrial respiratory pathways associated genes in the mutant HD2 iMac co-culture neuron and isogenic HD2 iMac co-culture NPCs population respectively. This suggests increased metabolic rates in these respective populations that are contributed to by the different variants of iMac during co-culture, and show that mutated iMicro can change the states of non-mutated isogenic organoid cells. A parallel process in vivo would likely have profound long-term consequences for brain development and homeostasis.

Altogether, we propose that CSF-1R mutation in HDLS may cause subtle yet impactful disruptions to microglia homeostasis due to abnormal metabolic functions in the developing fetal brain, that subsequently compound and result in aberrant microglia functions during adulthood. This could also be mediated by upregulated pro-inflammatory cytokine production such as IL-1β at a basal level or upon clearing of cellular debris during development and adulthood homeostasis. IL-1β is of particular interest as numerous studies reported the role of IL-1β in neuroinflammation and neurodegeneration especially within the field of Alzheimer’s disease: for example, postmortem brain sample analysis revealed elevated levels of Il-1β production, including around amyloid beta plaques; all suggestive of pathology-associated inflammation (35–37). Hence, we speculate that in CSF-1R+/- mutant microglia, which already have a predisposition towards increased secretion of IL-1β, are constantly being activated by IL-1β, either autocrine or paracrine, creating a dysfunctional neuro-environment that contributes to HDLS pathology. As such, experimental models looking to block or reduce IL-1β levels in the brain using the human mAb, canakinumab, targeting the proinflammatory cytokine IL-1β (38, 39), during both health and disease could be a potential target for therapy and may help to ameliorate HDLS pathology.

While no model perfectly recapitulates the full complexity of a disease, the application of iPSC-derived organoid co-culture approaches to primary patient-derived cells offers a valuable way to study the disease mechanisms of this rare pathology using human cells. Future studies employing microglia-sufficient patient-derived brain organoids have the potential to shed much-needed light into the pathogenesis of HDLS and other neurodegenerative diseases.